Variance reduction for MCMC methods via martingale representations

### 2019-12-11 aap/Revision/vr-mcmc-revision.tex
Revision

@@

\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. In this section   we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}\sum_{l=N+1}^{p}a_{p,l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\label{eq:29032018a5}\\
&=\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=N+1}^{N+n}\bar{a}_{l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and
\begin{align*}
\bar{a}_{l,\mathbf{k}}(x)
& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x) =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
\end{align*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we are interested in its conditional
Mean Squared Error (MSE),
which can be decomposed as the sum of the squared conditional bias
and the conditional variance:
\begin{eqnarray}\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)|\mathcal G_N\right]
&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2|\mathcal G_N\right]
\\
\nonumber
&=&\left(\mathsf{E}[\rho(f)|\mathcal G_N]-\pi(f)\right)^2
+\mathsf{Var}\left[\rho(f)|\mathcal G_N\right].
\end{eqnarray}
The quantities in~\eqref{eq:29032018a4}
are conditioned on $\mathcal G_N$,
as it reflects the way the estimators are used for MCMC:
the path of the Markov chain is simulated only once,
and we start to use the realized values
of the Markov chain to construct
our estimate only after the burn-in period of length~$N$.
We also notice that, due to the Markovian structure,
the conditioning on $\mathcal G_N$
in~\eqref{eq:29032018a4}
is equivalent to conditioning on $X_N$ only
(this is particularly clear in the case $\rho(f)=\pi_n^N(f)$
but requires some calculation
in the remaining case $\rho(f)=\pi_{K,n}^N(f)$).

\subsection{Squared conditional bias}
Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\[
\mathsf E\left[\left.M_{K,n}^N(f)\right|\mathcal G_N\right]=0,
\]
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
Notice that this remains true also when we substitute
the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$ For a bounded Borel function~$f$,
we can estimate the conditional bias
similarly to the beginning of ~\cite[Section~4]{durmus:moulines:2017}:
\begin{align}
\left(\mathsf{E}[\pi_{K,n}^N(f)|\mathcal G_N]-\pi(f)\right)^2
&=\left(\mathsf{E}[\pi_{n}^N(f)|\mathcal G_N]-\pi(f)\right)^2
\label{eq:06042018a2}\\
&\le
\mathrm{osc}(f)^2
\sum_{p=N+1}^{N+n}
\omega_{p,n}^N\,
\|Q_\gamma^{N+1,p}(X_N,\cdot)-\pi(\cdot)\|_{TV}^2,
\notag
\end{align}
where
$\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$,
$\|\mu-\nu\|_{\TV}$ denotes the total variation distance
between probability measures $\mu$ and $\nu$, that is,
$$
\|\mu-\nu\|_{TV}=\sup_{A\in\mathcal B(\mathbb R^d)}
|\mu(A)-\nu(A)|,
$$
$\pi(\cdot)$ denotes the probability measure on $\mathbb R^d$
with density $\pi$ of~\eqref{eq:stationary_distr};
for $\gamma>0$, the Markov kernel $R_\gamma$
from $(\mathbb R^d,\mathcal B(\mathbb R^d))$
into $(\mathbb R^d,\mathcal B(\mathbb R^d))$
is defined by
$$
R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,
$$
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_\gamma^{k,l}$ is
$
Q_\gamma^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k},
$
which, finally, provides the (random) measure
$Q_\gamma^{N+1,p}(X_N,\cdot)$
used in the right-hand side of~\eqref{eq:06042018a2}.
Different specific upper bounds for the squared bias
can be deduced from~\eqref{eq:06042018a2}
using results of Section~3 in~\cite{durmus:moulines:2017}
on bounds in the total variation distance.

\subsection{Conditional variance}
An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
is provided in  in~\cite[Theorem~17]{durmus:moulines:2017}.
As for the estimator~\eqref{eq:29032018a3}, we introduce lemma of conditional variance decomposition.
It follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
\begin{lem}
\label{cond_var_dec}
	For variance-reduced estimator~\eqref{eq:29032018a3}, defined for the Markov Chain~\eqref{eq:chain_gen} the following representation holds
\begin{equation}\label{eq:29032018a6}
\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]
%=\sum_{\min_i (k_i)\ge K+1}
=\sum_{\mathbf k\colon \|\mathbf k\|\ge K+1}
\sum_{l=N+1}^{N+n}
\mathsf{E}\left[\left.\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}
\right|\mathcal G_N\right].
\end{equation}
\end{lem}
Now we present an upper bound for the right-hand sides of~\eqref{eq:29032018a6} in case of ULA algorithm.
\begin{thm}\label{th:mr}
Fix $K\in\mathbb N$.
Suppose that $f$  and \(\mu\) are \(K+1\) times continuously differentiable
and it holds
\begin{eqnarray*}
|\partial^{\mathbf{k}} f(x)|\le  B_f,\quad |\partial^{\mathbf{k}}\mu(x)|\le  B_\mu, \quad
x\in\mathbb R^d
\end{eqnarray*}
for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  K+1\),
\[
\jac{\mu}(x)\ge b_\mu I ,\quad x\in\mathbb R^d,
\]
for some positive number $b_\mu\in(0,B_\mu]$  where  \(\jac{\mu}\) stands for Jacobian of \(\mu.\)
Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$.
We  assume that $\gamma_1<\frac1{B_\mu}$ and that
\begin{equation}\label{eq:17042018a1}
\sum_{r=j}^\infty \gamma_r\prod_{k=j}^{r}\left[1-\gamma_{k}b_\mu\right]\leq C,\quad\text{for all }j\in\mathbb N,
\end{equation}
with some constant \(C\). Then it holds
\begin{eqnarray}
\label{eq:var-bound}
\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma_{l}}{2}\right)^{|I|K},
\end{eqnarray}
where the sum in \eqref{eq:var-bound} is taken over all subsets \(I\) of \(\{1,\ldots,d\}\) and \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)
\end{thm}
\begin{remark}
Note that in  Theorem~\eqref{th:mr} we don't require that  the function \(\mu\) itself is bounded.
Assumption~\eqref{eq:17042018a1} is not restrictive.
For instance, a straightforward calculation shows
that \eqref{eq:17042018a1} is satisfied in most interesting case
$\gamma_k=\const/k^\alpha$ with $\alpha\in (0,1)$.
\end{remark}

@@

\begin{thm}
Suppose that
\begin{eqnarray*}
\sum_{p=1}^{\infty} \gamma^2_p<\infty.
\end{eqnarray*}
Under assumptions (H1) and (H2), then for any bounded function $f$ on $\mathbb{R}^d$ 
we have
\begin{eqnarray*}
\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}},
\end{eqnarray*}
 and
\begin{eqnarray*}
\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
\end{eqnarray*}
Moreover 
\begin{eqnarray*}
\mathsf{E}\left[\exp\left(\rho\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
\end{eqnarray*}
for some small enough $\rho>0,$  all natural $0<l\leq p<\infty$ and constant $C$ not depending on $l,p.$
\end{thm}

>>>>

\par

@@

The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
should be small for \(K\) large enough. In this section   we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}

We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}\sum_{l=N+1}^{p}a_{p,l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\label{eq:29032018a5}\\
&=\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=N+1}^{N+n}\bar{a}_{l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and
\begin{align*}
\bar{a}_{l,\mathbf{k}}(x)
& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x) =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
\end{align*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{eqnarray}\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right]
&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2\right]
\\
\nonumber
&=&\left(\mathsf{E}[\rho(f)]-\pi(f)\right)^2
+\mathsf{Var}[\rho(f)].
\end{eqnarray}
Our analysis will be done under the following two assumptions.

\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L\geq 0$ such that 
\begin{eqnarray*}
\|\nabla U(x)-\nabla U(y)\|\leq L\|x-y\|, \quad x,y\in \mathbb{R}^d.
\end{eqnarray*}

\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)	$ and any $y\in \mathbb{R}^d$ it holds 
\langle D^2U(x)y,y\rangle \geq m\|y\|^2, \quad \|D^3 U(y)\| \leq M. 
First we analyse the biases of the estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
\subsection{Squared bias}
Due to the martingale transform structure
of $M_{K,n}^N(f)$,
\mathsf E\left[M_{K,n}^N(f)\right]=0,
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
Notice that this remains true also when we substitute
the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$ For a bounded Borel function~$f$,
we can estimate the conditional bias
similarly to the beginning of ~\cite[Section~4]{durmus:moulines:2017}:
\left(\mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
&=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
\label{eq:06042018a2}\\
\mathrm{osc}(f)^2
\sum_{p=N+1}^{N+n}
\omega_{p,n}^N\,
\|Q_\gamma^{p}(x_0,\cdot)-\pi(\cdot)\|_{TV}^2,
$\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$,
$\|\mu-\nu\|_{\TV}$ denotes the total variation distance
between probability measures $\mu$ and $\nu$, that is,
\|\mu-\nu\|_{TV}=\sup_{A\in\mathcal B(\mathbb R^d)}
|\mu(A)-\nu(A)|,
$\pi(\cdot)$ denotes the probability measure on $\mathbb R^d$
with density $\pi$ of~\eqref{eq:stationary_distr};
for $\gamma>0$, the Markov kernel $R_\gamma$
from $(\mathbb R^d,\mathcal B(\mathbb R^d))$
into $(\mathbb R^d,\mathcal B(\mathbb R^d))$
is defined by
R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_\gamma^{k,l}$ is
$
Q_\gamma^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k},
$
which, finally, provides the (random) measure
$Q_\gamma^{N+1,p}(X_N,\cdot)$
used in the right-hand side of~\eqref{eq:06042018a2}.
Different specific upper bounds for the squared bias
can be deduced from~\eqref{eq:06042018a2}
using results of Section~3 in~\cite{durmus:moulines:2017}
on bounds in the total variation distance. For example, the following proposition holds.
Suppose that the assumption (H1) holds, then it holds
\|Q_\gamma^{p}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \sum_{k=r+1}^{p} \gamma_k^2+\exp\left(-\varkappa\sum_{k=r+1}^{p} \gamma_k\right),
for some $\varkappa>0,$ where $\lesssim$ stands for  inequality up to a constant depending on $L$ and $\mathcal G_N.$ In particular if $\gamma_k=k^{-\alpha}$ for some $\alpha\in (0.5,1], $ then 
\|Q_\gamma^{N+1,N+n}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \frac{1}{(N+1)^{2\alpha-1}}
provided that $N/n\to 0$ as $N,n\to \infty.$
\subsection{Conditional variance}
An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
is provided in  in~\cite[Theorem~17]{durmus:moulines:2017}.
As for the estimator~\eqref{eq:29032018a3}, we introduce lemma of conditional variance decomposition.
It follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
\label{cond_var_dec}
	For variance-reduced estimator~\eqref{eq:29032018a3}, defined for the Markov Chain~\eqref{eq:chain_gen} the following representation holds
\begin{equation}\label{eq:29032018a6}
\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]
%=\sum_{\min_i (k_i)\ge K+1}
=\sum_{\mathbf k\colon \|\mathbf k\|\ge K+1}
\sum_{l=N+1}^{N+n}
\mathsf{E}\left[\left.\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}
\right|\mathcal G_N\right].
Now we present an upper bound for the right-hand sides of~\eqref{eq:29032018a6} in case of ULA algorithm.
\begin{thm}\label{th:mr}
Suppose that $f$  is two times continuously differentiable
and it holds
|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad
x\in\mathbb R^d
for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Moreover assume that there exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)$ and any $y\in \mathbb{R}^d$ it holds
\langle \nabla \mu(x)y,y\rangle \geq m\|y\|^2.
Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$ and $\sum_{k=1}^\infty \gamma^2_k<\infty.$
Then it holds under 
\label{eq:var-bound}
\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},
where \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)

@@

\begin{thm}
Suppose that
\begin{eqnarray*}
\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
\end{eqnarray*}
Under assumptions (H1) and (H2), then for any bounded function $g$ on $\mathbb{R}^d$ 
we have
\begin{eqnarray*}
\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
\end{eqnarray*}
 and
\begin{eqnarray*}
\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
\end{eqnarray*}
Moreover, 
\begin{eqnarray*}
\mathsf{E}\left[\exp\left(\rho\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
\end{eqnarray*}
for some small enough $\rho>0,$  all natural $0<l\leq p<\infty$ and constant $C$ not depending on $l,p.$
\end{thm}
